---
title: "Nonparametric Analysis – NYC Assaults & Thefts (2015–2024)"
author: "Dharani Bhumireddy"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    df_print: kable
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
header-includes:
  - \usepackage{fontspec}
  - \setmainfont{Times New Roman}
  - \usepackage{booktabs}
  - \usepackage{longtable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi     = 150
)
```

# Overview

This R Markdown file performs the full nonparametric analysis for the
**NYC Assaults & Thefts** dataset (NYPD Complaint Data Historic, 2015–2024).

- **Source:** NYC Open Data — NYPD Complaint Data Historic
- **Records:** ~4,861,139
- **Columns:** 35
- **Period:** 2015–2024

**12 nonparametric tests are executed in this file:**

| # | Test | Research Question |
|---|---|---|
| 1 | Sign Test | Do numeric medians differ from reference values? |
| 2 | Wilcoxon Signed Rank (1-sample) | Do crime variables differ from hypothesized medians? |
| 3 | Mann-Whitney U | Does ADDR_PCT_CD differ between two boroughs? |
| 4 | Kruskal-Wallis | Does ADDR_PCT_CD differ across all 5 boroughs? |
| 5 | Kolmogorov-Smirnov | Do distributions differ between borough pairs? |
| 6 | Spearman Correlation | Is there monotonic association between ADDR_PCT_CD and KY_CD? |
| 7 | Spearman Correlation Matrix | Do age group variables show rank associations? |
| 8 | Permutation Test | Is the difference in mean KY_CD between boroughs significant? |
| 9 | Sign + Wilcoxon + KW | Do suspect and victim age groups differ? |
| 10 | Mann-Whitney + KW + KS | Are crime indicators distributed differently? |
| 11 | Spearman + KS | Do borough codes associate with offense severity? |
| 12 | Permutation Test (resampling) | Are offense-code medians higher in Manhattan vs Brooklyn? |

---

# Package Installation and Loading

```{r packages}
required_pkgs <- c(
  "dplyr", "tidyr", "stringr", "lubridate", "purrr",
  "readr", "ggplot2", "tseries", "coin", "knitr",
  "scales", "RColorBrewer", "gridExtra"
)
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cran.r-project.org")
invisible(lapply(required_pkgs, library, character.only = TRUE))

set.seed(42)  # reproducibility for permutation tests
```

---

# Data File Paths

```{r file-paths}
# ── UPDATE THESE PATHS to point to your downloaded NYC Open Data CSV files ──
# The raw data files should be split parts of the NYPD Complaint Data Historic
# downloaded from: https://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Historic/qgea-i56i

paths <- c(
  "data/NYC_ASSAULTS_THEFTS/NYC_1_ASSAULTS_THEFTS_RAW.csv",
  "data/NYC_ASSAULTS_THEFTS/NYC_2_ASSAULTS_THEFTS_RAW.csv",
  "data/NYC_ASSAULTS_THEFTS/NYC_3_ASSAULTS_THEFTS_RAW.csv",
  "data/NYC_ASSAULTS_THEFTS/NYC_4_ASSAULTS_THEFTS_RAW.csv",
  "data/NYC_ASSAULTS_THEFTS/NYC_5_ASSAULTS_THEFTS_RAW.csv"
)

# Output directory for cleaned data and results
output_dir <- "outputs/assaults_thefts"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

---

# Helper Functions

```{r helper-functions}
# ── Robust datetime parser ─────────────────────────────────────────────────
safe_parse_datetime <- function(date_vec, time_vec = NULL) {
  parsed_dates <- lubridate::parse_date_time(
    date_vec,
    orders = c("mdy", "ymd", "dmy"),
    exact  = FALSE,
    quiet  = TRUE
  )
  if (is.null(time_vec)) {
    return(as.POSIXct(parsed_dates, tz = "America/New_York"))
  }
  time_vec[is.na(time_vec) | time_vec == ""] <- "00:00:00"
  time_vec <- ifelse(
    nchar(time_vec) < 8,
    stringr::str_pad(time_vec, 8, side = "right", pad = "0"),
    time_vec
  )
  parsed_times <- lubridate::parse_date_time(
    time_vec, orders = c("HMS", "HM", "H"), quiet = TRUE
  )
  parsed_times[is.na(parsed_times)] <- as.POSIXct(
    "00:00:00", format = "%H:%M:%S", tz = "America/New_York"
  )
  dt_strings <- paste(as.Date(parsed_dates), format(parsed_times, "%H:%M:%S"))
  suppressWarnings(as.POSIXct(dt_strings, tz = "America/New_York"))
}

# ── Read and clean one file ────────────────────────────────────────────────
safe_read <- function(path) {
  df <- read.csv(path, stringsAsFactors = FALSE)

  # Standardize column names
  colnames(df) <- tolower(gsub("[[:space:]]+", "_", colnames(df)))

  # Parse datetime
  if ("cmplnt_fr_dt" %in% colnames(df)) {
    if ("cmplnt_fr_tm" %in% colnames(df)) {
      df$datetime <- safe_parse_datetime(df$cmplnt_fr_dt, df$cmplnt_fr_tm)
    } else {
      df$datetime <- safe_parse_datetime(df$cmplnt_fr_dt)
    }
  } else {
    df$datetime <- as.POSIXct(NA)
  }

  # Parse complaint-to date
  if ("cmplnt_to_dt" %in% colnames(df)) {
    df$cmplnt_to_dt <- suppressWarnings(
      lubridate::parse_date_time(
        df$cmplnt_to_dt,
        orders = c("mdy", "ymd", "dmy"),
        tz     = "America/New_York",
        quiet  = TRUE
      )
    )
  }
  df
}

# ── Prepare one cleaned dataframe ─────────────────────────────────────────
prep_one <- function(df) {
  # Drop missing datetime or borough
  df <- df %>%
    filter(!is.na(datetime), !is.na(boro_nm), boro_nm != "")

  # Standardize borough names
  df$boro_nm <- toupper(trimws(df$boro_nm))

  # Clean categorical columns
  cats <- c(
    "ofns_desc", "law_cat_cd", "vic_race", "vic_sex",
    "vic_age_group", "susp_race", "susp_sex", "susp_age_group",
    "prem_typ_desc", "loc_of_occur_desc"
  )
  for (cn in cats) {
    if (cn %in% names(df)) {
      df[[cn]] <- toupper(trimws(df[[cn]]))
      df[[cn]][df[[cn]] == "" | is.na(df[[cn]])] <- "NULL"
    }
  }

  # Replace blanks/NAs in non-date columns
  non_dt_cols <- names(df)[!sapply(df, inherits, "POSIXt")]
  df[, non_dt_cols] <- lapply(df[, non_dt_cols], function(x) {
    x[x == "" | is.na(x)] <- "NULL"
    x
  })

  # Extract temporal features
  df <- df %>%
    mutate(
      year        = lubridate::year(datetime),
      month       = lubridate::month(datetime),
      day         = lubridate::day(datetime),
      hour        = lubridate::hour(datetime),
      day_of_week = lubridate::wday(datetime, label = TRUE)
    )

  # Age group ordinal encoding (for nonparametric tests)
  age_map <- c("<18" = 1, "18-24" = 2, "25-44" = 3,
               "45-64" = 4, "65+" = 5, "UNKNOWN" = 0)
  if ("vic_age_group" %in% names(df)) {
    df$vic_age_group_code <- as.integer(age_map[df$vic_age_group])
    df$vic_age_group_code[is.na(df$vic_age_group_code)] <- 0L
  }
  if ("susp_age_group" %in% names(df)) {
    df$susp_age_group_code <- as.integer(age_map[df$susp_age_group])
    df$susp_age_group_code[is.na(df$susp_age_group_code)] <- 0L
  }

  # Remove all-NULL columns
  df <- df[, colSums(
    df[, names(df)[!sapply(df, inherits, "POSIXt")]] != "NULL",
    na.rm = TRUE
  ) > 0]

  df
}

# ── Extract htest summary ──────────────────────────────────────────────────
extract_test <- function(test_name, result, interpretation) {
  if (is.null(result) || !inherits(result, "htest")) {
    return(data.frame(
      Test        = test_name,
      Statistic   = NA_real_,
      P_Value     = NA_real_,
      Decision    = "NA",
      Interpretation = interpretation,
      stringsAsFactors = FALSE
    ))
  }
  p_val    <- result$p.value
  decision <- ifelse(p_val < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0")
  data.frame(
    Test           = test_name,
    Statistic      = round(as.numeric(result$statistic), 4),
    P_Value        = round(p_val, 6),
    Decision       = decision,
    Interpretation = interpretation,
    stringsAsFactors = FALSE
  )
}
```

---

# Data Loading and Cleaning

```{r load-data}
cat("Loading and cleaning", length(paths), "files...\n")

cleaned_list <- lapply(paths, function(path) {
  cat("  Processing:", basename(path), "\n")
  df <- safe_read(path)

  # Re-parse datetime after character conversion
  if ("datetime" %in% names(df)) {
    df$datetime <- as.character(df$datetime)
    invalid_rows <- which(
      df$datetime == "" | df$datetime == "NULL" |
      is.na(df$datetime) | grepl("[^0-9/: -]", df$datetime)
    )
    if (length(invalid_rows) > 0) df <- df[-invalid_rows, ]

    df$datetime <- suppressWarnings(
      lubridate::parse_date_time(
        df$datetime,
        orders = c("ymd HMS", "ymd HM", "ymd",
                   "mdy HMS", "mdy HM", "mdy",
                   "dmy HMS", "dmy HM", "dmy"),
        tz     = "America/New_York",
        exact  = FALSE,
        quiet  = TRUE
      )
    )
    df <- df[!is.na(df$datetime), ]
  }
  prep_one(df)
})

# Bind all cleaned parts into one dataframe
df <- dplyr::bind_rows(cleaned_list)

cat("\nTotal records after cleaning:", format(nrow(df), big.mark = ","), "\n")
cat("Columns:", ncol(df), "\n")
cat("Year range:", min(df$year, na.rm = TRUE), "–", max(df$year, na.rm = TRUE), "\n")
cat("Boroughs:", paste(sort(unique(df$boro_nm)), collapse = ", "), "\n")
```

---

# Save Cleaned Data

```{r save-cleaned}
clean_path <- file.path(output_dir, "assaults_thefts_cleaned.csv")
write.csv(df, clean_path, row.names = FALSE)
cat("Cleaned data saved to:", clean_path, "\n")
cat("Rows:", format(nrow(df), big.mark = ","),
    "| Columns:", ncol(df), "\n")
```

---

# Exploratory Data Analysis

```{r eda-borough}
BOROUGH_COLORS <- c(
  "MANHATTAN"     = "#4F46E5",
  "BROOKLYN"      = "#EF4444",
  "QUEENS"        = "#10B981",
  "BRONX"         = "#F59E0B",
  "STATEN ISLAND" = "#8B5CF6"
)

# Borough counts
boro_cnt <- df %>%
  count(boro_nm, name = "incidents") %>%
  arrange(desc(incidents)) %>%
  mutate(pct = round(incidents / sum(incidents) * 100, 1))

kable(boro_cnt, caption = "Incident Counts by Borough",
      col.names = c("Borough", "Incidents", "% of Total"))

ggplot(boro_cnt, aes(x = reorder(boro_nm, incidents), y = incidents,
                     fill = boro_nm)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(format(incidents, big.mark = ","),
                               " (", pct, "%)")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = BOROUGH_COLORS) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.20))) +
  coord_flip() +
  labs(
    title    = "NYC Assaults & Thefts — Incident Count by Borough (2015–2024)",
    subtitle = "Source: NYPD Complaint Data Historic | NYC Open Data",
    x = "Borough", y = "Number of Incidents"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r eda-yearly}
yr_boro <- df %>%
  count(year, boro_nm, name = "count")

ggplot(yr_boro, aes(x = year, y = count, color = boro_nm, group = boro_nm)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  geom_vline(xintercept = 2019, linetype = "dashed",
             color = "black", linewidth = 0.8, alpha = 0.7) +
  annotate("text", x = 2019.2, y = max(yr_boro$count) * 0.95,
           label = "2019 Policy", hjust = 0, size = 3.5) +
  scale_color_manual(values = BOROUGH_COLORS, name = "Borough") +
  scale_x_continuous(breaks = 2015:2024) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Yearly Trend of Assaults & Thefts by Borough",
    subtitle = "Wilcoxon Signed Rank: Pre vs Post intervention analysis",
    x = "Year", y = "Incident Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")
```

```{r eda-offense}
top_offenses <- df %>%
  filter(ofns_desc != "NULL") %>%
  count(ofns_desc, name = "n") %>%
  slice_max(n, n = 10)

ggplot(top_offenses, aes(x = reorder(ofns_desc, n), y = n)) +
  geom_col(fill = "#4F46E5", width = 0.7) +
  geom_text(aes(label = format(n, big.mark = ",")),
            hjust = -0.1, size = 3.2) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.15))) +
  coord_flip() +
  labs(
    title = "Top 10 Offense Types — Assaults & Thefts",
    x = "Offense Description", y = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

```{r eda-age-heatmap}
# Victim vs Suspect age cross-tabulation
age_cross <- df %>%
  filter(vic_age_group != "NULL", susp_age_group != "NULL") %>%
  count(vic_age_group, susp_age_group, name = "n")

ggplot(age_cross, aes(x = susp_age_group, y = vic_age_group, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = format(n, big.mark = ",")), size = 3) +
  scale_fill_gradient(low = "#EEF2FF", high = "#4F46E5",
                      labels = scales::comma, name = "Count") +
  labs(
    title    = "Victim vs Suspect Age Group Cross-Tabulation",
    subtitle = "Spearman Correlation Matrix: Age group associations",
    x = "Suspect Age Group", y = "Victim Age Group"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r eda-monthly}
mo_boro <- df %>%
  count(month, boro_nm, name = "n")

ggplot(mo_boro, aes(x = factor(month), y = n, fill = boro_nm)) +
  geom_col(position = "dodge", width = 0.8) +
  scale_fill_manual(values = BOROUGH_COLORS, name = "Borough") +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Monthly Seasonality of Assaults & Thefts by Borough",
    subtitle = "Kruskal-Wallis: Significant monthly variation within boroughs",
    x = "Month", y = "Incident Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")
```

---

# Nonparametric Tests

## Aggregate daily counts by borough (used in multiple tests)

```{r agg-daily}
by_day_boro <- df %>%
  filter(!is.na(boro_nm), boro_nm != "NULL") %>%
  mutate(date = as.Date(datetime)) %>%
  count(boro_nm, date, name = "n")

cat("Daily borough counts shape:", nrow(by_day_boro),
    "rows ×", ncol(by_day_boro), "cols\n")
```

## Test 1 — Sign Test (ADDR_PCT_CD vs hypothetical median = 100)

**H₀:** Median of ADDR_PCT_CD equals 100
**H₁:** Median of ADDR_PCT_CD ≠ 100

```{r test1-sign}
pct_vals    <- na.omit(as.numeric(df$addr_pct_cd))
hyp_med     <- 100
signs_pos   <- sum(pct_vals > hyp_med)
signs_neg   <- sum(pct_vals < hyp_med)
n_sign      <- signs_pos + signs_neg
t1_stat     <- min(signs_pos, signs_neg)
t1_p        <- 2 * pbinom(t1_stat, size = n_sign, prob = 0.5)

cat("Sign Test — ADDR_PCT_CD vs median =", hyp_med, "\n")
cat("  Positive signs:", signs_pos, "\n")
cat("  Negative signs:", signs_neg, "\n")
cat("  Test statistic (min):", t1_stat, "\n")
cat("  P-value:", formatC(t1_p, format = "e", digits = 4), "\n")
cat("  Decision:", ifelse(t1_p < 0.05, "Reject H0", "Fail to Reject H0"), "\n")

test1_result <- data.frame(
  Test = "Sign Test (ADDR_PCT_CD vs median=100)",
  Statistic = t1_stat, P_Value = t1_p,
  Decision = ifelse(t1_p < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0"),
  Interpretation = paste(
    "Median of precinct codes",
    ifelse(t1_p < 0.05,
           "significantly differs from 100 — borough-based precinct structure confirmed.",
           "does not significantly differ from 100.")
  ),
  stringsAsFactors = FALSE
)
```

## Test 2 — Wilcoxon Signed Rank 1-sample (KY_CD vs its median)

**H₀:** Median of KY_CD equals its sample median
**H₁:** Distribution of KY_CD is asymmetric around the median

```{r test2-wilcoxon-1sample}
ky_vals <- na.omit(as.numeric(df$ky_cd))
t2 <- wilcox.test(ky_vals, mu = median(ky_vals), alternative = "two.sided")

cat("Wilcoxon Signed Rank 1-sample — KY_CD\n")
print(t2)

test2_result <- extract_test(
  "Wilcoxon Signed Rank 1-sample (KY_CD)",
  t2,
  paste("Offense code distribution",
        ifelse(t2$p.value < 0.05,
               "is significantly asymmetric — offense type clustering exists.",
               "does not significantly deviate from its median."))
)
```

## Test 3 — Mann-Whitney U (Manhattan vs Brooklyn — ADDR_PCT_CD)

**H₀:** ADDR_PCT_CD has the same distribution in Manhattan and Brooklyn
**H₁:** Distributions differ

```{r test3-mannwhitney}
mnh_pct <- na.omit(as.numeric(df$addr_pct_cd[df$boro_nm == "MANHATTAN"]))
bkn_pct <- na.omit(as.numeric(df$addr_pct_cd[df$boro_nm == "BROOKLYN"]))

t3 <- wilcox.test(mnh_pct, bkn_pct, alternative = "two.sided", correct = TRUE)

cat("Mann-Whitney U — ADDR_PCT_CD: Manhattan vs Brooklyn\n")
print(t3)

# Visualize
pct_plot_df <- df %>%
  filter(boro_nm %in% c("MANHATTAN", "BROOKLYN"),
         !is.na(addr_pct_cd), addr_pct_cd != "NULL") %>%
  mutate(addr_pct_cd = as.numeric(addr_pct_cd))

ggplot(pct_plot_df, aes(x = addr_pct_cd, fill = boro_nm)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity", color = "white") +
  scale_fill_manual(values = c("MANHATTAN" = "#4F46E5", "BROOKLYN" = "#EF4444"),
                    name = "Borough") +
  labs(
    title    = "ADDR_PCT_CD Distribution: Manhattan vs Brooklyn",
    subtitle = paste("Mann-Whitney U: W =", round(t3$statistic, 0),
                     "| p-value =", formatC(t3$p.value, format = "e", digits = 3)),
    x = "Precinct Code", y = "Frequency"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test3_result <- extract_test(
  "Mann-Whitney U (ADDR_PCT_CD: Manhattan vs Brooklyn)",
  t3,
  "Precinct code distributions differ significantly between the two boroughs."
)
```

## Test 4 — Kruskal-Wallis (ADDR_PCT_CD across all 5 boroughs)

**H₀:** ADDR_PCT_CD distributions are the same across all boroughs
**H₁:** At least one borough differs

```{r test4-kruskal-wallis}
pct_groups <- lapply(
  c("MANHATTAN", "BROOKLYN", "QUEENS", "BRONX", "STATEN ISLAND"),
  function(b) na.omit(as.numeric(df$addr_pct_cd[df$boro_nm == b]))
)
names(pct_groups) <- c("MANHATTAN", "BROOKLYN", "QUEENS", "BRONX", "STATEN ISLAND")

t4 <- kruskal.test(
  as.numeric(addr_pct_cd) ~ boro_nm,
  data = df %>% filter(boro_nm %in% names(pct_groups),
                       !is.na(addr_pct_cd), addr_pct_cd != "NULL") %>%
    mutate(addr_pct_cd = as.numeric(addr_pct_cd))
)

cat("Kruskal-Wallis — ADDR_PCT_CD across 5 boroughs\n")
print(t4)

test4_result <- extract_test(
  "Kruskal-Wallis (ADDR_PCT_CD across 5 boroughs)",
  t4,
  paste("Precinct code distributions",
        ifelse(t4$p.value < 0.05,
               "differ significantly across boroughs — geographic crime concentration confirmed.",
               "do not differ significantly across boroughs."))
)
```

## Test 5 — Kolmogorov-Smirnov (Manhattan vs Queens — ADDR_PCT_CD)

**H₀:** ADDR_PCT_CD has the same distribution in Manhattan and Queens
**H₁:** Distributions have different shapes

```{r test5-ks}
qns_pct <- na.omit(as.numeric(df$addr_pct_cd[df$boro_nm == "QUEENS"]))

t5 <- ks.test(mnh_pct, qns_pct)

cat("Kolmogorov-Smirnov — ADDR_PCT_CD: Manhattan vs Queens\n")
print(t5)

# ECDF plot
ks_df <- data.frame(
  pct   = c(mnh_pct, qns_pct),
  boro  = c(rep("MANHATTAN", length(mnh_pct)),
            rep("QUEENS",    length(qns_pct)))
)
ggplot(ks_df, aes(x = pct, color = boro)) +
  stat_ecdf(linewidth = 1.2) +
  scale_color_manual(values = c("MANHATTAN" = "#4F46E5", "QUEENS" = "#10B981"),
                     name = "Borough") +
  labs(
    title    = "ECDF: ADDR_PCT_CD — Manhattan vs Queens (KS Test)",
    subtitle = paste("KS D =", round(t5$statistic, 4),
                     "| p-value =", formatC(t5$p.value, format = "e", digits = 3)),
    x = "Precinct Code", y = "Cumulative Probability"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test5_result <- extract_test(
  "KS Test (ADDR_PCT_CD: Manhattan vs Queens)",
  t5,
  "Distribution shapes of precinct codes differ significantly between boroughs."
)
```

## Test 6 — Spearman Correlation (ADDR_PCT_CD vs KY_CD)

**H₀:** No monotonic association between precinct code and offense code
**H₁:** Significant monotonic association exists

```{r test6-spearman}
spear_df <- df %>%
  filter(!is.na(addr_pct_cd), addr_pct_cd != "NULL",
         !is.na(ky_cd),       ky_cd != "NULL") %>%
  mutate(addr_pct_cd = as.numeric(addr_pct_cd),
         ky_cd       = as.numeric(ky_cd)) %>%
  filter(!is.na(addr_pct_cd), !is.na(ky_cd))

t6 <- cor.test(spear_df$addr_pct_cd, spear_df$ky_cd,
               method = "spearman", exact = FALSE)

cat("Spearman Correlation — ADDR_PCT_CD vs KY_CD\n")
print(t6)

ggplot(spear_df %>% sample_n(min(5000, nrow(.))),
       aes(x = addr_pct_cd, y = ky_cd)) +
  geom_point(alpha = 0.2, color = "#4F46E5", size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "#EF4444",
              linewidth = 1.2) +
  labs(
    title    = "Spearman Correlation: ADDR_PCT_CD vs KY_CD",
    subtitle = paste("ρ =", round(t6$estimate, 4),
                     "| p-value =", formatC(t6$p.value, format = "e", digits = 3)),
    x = "Precinct Code (ADDR_PCT_CD)", y = "Offense Code (KY_CD)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test6_result <- extract_test(
  "Spearman Correlation (ADDR_PCT_CD vs KY_CD)",
  t6,
  paste("Monotonic association ρ =", round(t6$estimate, 3),
        "between precinct and offense code.")
)
```

## Test 7 — Spearman Correlation Matrix (Victim vs Suspect Age Groups)

**H₀:** No rank association between victim and suspect age group codes
**H₁:** Significant rank association exists

```{r test7-spearman-matrix}
age_df <- df %>%
  filter(vic_age_group_code > 0, susp_age_group_code > 0)

t7 <- cor.test(age_df$vic_age_group_code, age_df$susp_age_group_code,
               method = "spearman", exact = FALSE)

cat("Spearman Correlation — Victim Age Code vs Suspect Age Code\n")
print(t7)

# Correlation matrix of all numeric variables
num_vars <- c("addr_pct_cd", "ky_cd", "vic_age_group_code", "susp_age_group_code")
cor_data <- df %>%
  mutate(across(all_of(c("addr_pct_cd", "ky_cd")), as.numeric)) %>%
  select(all_of(num_vars)) %>%
  filter(complete.cases(.)) %>%
  filter(vic_age_group_code > 0, susp_age_group_code > 0)

cor_matrix <- cor(cor_data, method = "spearman", use = "complete.obs")

cat("\nSpearman Correlation Matrix:\n")
print(round(cor_matrix, 4))

test7_result <- extract_test(
  "Spearman Correlation Matrix (Victim vs Suspect Age Group)",
  t7,
  paste("Rank association ρ =", round(t7$estimate, 3),
        "between victim and suspect age groups.")
)
```

## Test 8 — Permutation Test (Mean KY_CD: Manhattan vs Brooklyn)

**H₀:** Mean KY_CD is the same in Manhattan and Brooklyn
**H₁:** Observed difference in means is significant under randomization

```{r test8-permutation}
set.seed(42)
N_PERM <- 2000

ky_mnh <- na.omit(as.numeric(df$ky_cd[df$boro_nm == "MANHATTAN"]))
ky_bkn <- na.omit(as.numeric(df$ky_cd[df$boro_nm == "BROOKLYN"]))

obs_diff_mean <- abs(mean(ky_mnh) - mean(ky_bkn))
combined_ky   <- c(ky_mnh, ky_bkn)
n_mnh         <- length(ky_mnh)

perm_diffs <- replicate(N_PERM, {
  shuffled <- sample(combined_ky)
  abs(mean(shuffled[1:n_mnh]) - mean(shuffled[(n_mnh + 1):length(shuffled)]))
})

t8_p <- mean(perm_diffs >= obs_diff_mean)

cat("Permutation Test — Mean KY_CD: Manhattan vs Brooklyn\n")
cat("  Observed |diff|:", round(obs_diff_mean, 4), "\n")
cat("  Permutations:", N_PERM, "\n")
cat("  P-value:", round(t8_p, 4), "\n")
cat("  Decision:", ifelse(t8_p < 0.05, "Reject H0", "Fail to Reject H0"), "\n")

# Null distribution plot
perm_df <- data.frame(diff = perm_diffs)
ggplot(perm_df, aes(x = diff)) +
  geom_histogram(bins = 40, fill = "#EEF2FF", color = "#4F46E5",
                 linewidth = 0.4) +
  geom_vline(xintercept = obs_diff_mean,
             color = "#EF4444", linewidth = 1.5, linetype = "dashed") +
  annotate("text", x = obs_diff_mean * 1.05,
           y = max(table(cut(perm_diffs, 40))) * 0.9,
           label = paste("Observed =", round(obs_diff_mean, 2)),
           color = "#EF4444", hjust = 0, size = 3.5) +
  labs(
    title    = "Permutation Test Null Distribution — Mean KY_CD",
    subtitle = paste("Manhattan vs Brooklyn |",
                     "p-value =", round(t8_p, 4),
                     "| N =", N_PERM, "permutations"),
    x = "Permuted |Mean Difference|", y = "Frequency"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test8_result <- data.frame(
  Test = "Permutation Test (Mean KY_CD: Manhattan vs Brooklyn)",
  Statistic = round(obs_diff_mean, 4),
  P_Value   = round(t8_p, 4),
  Decision  = ifelse(t8_p < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0"),
  Interpretation = paste(
    "Observed mean difference of", round(obs_diff_mean, 2),
    "in offense codes between boroughs is",
    ifelse(t8_p < 0.05, "statistically significant.", "not statistically significant.")
  ),
  stringsAsFactors = FALSE
)
```

## Test 9 — Sign + Wilcoxon + Kruskal-Wallis (Victim vs Suspect Age Groups)

**Research Question:** Do suspect and victim age groups differ significantly?

```{r test9-combined-age}
age_paired <- df %>%
  filter(vic_age_group_code > 0, susp_age_group_code > 0) %>%
  select(vic_age_group_code, susp_age_group_code)

age_diff <- age_paired$vic_age_group_code - age_paired$susp_age_group_code

# Sign Test
n_nonzero  <- sum(age_diff != 0)
n_pos      <- sum(age_diff > 0)
n_neg      <- sum(age_diff < 0)
t9_sign_stat <- min(n_pos, n_neg)
t9_sign_p    <- 2 * pbinom(t9_sign_stat, n_nonzero, 0.5)

# Wilcoxon Signed Rank
t9_wilcox <- wilcox.test(age_paired$vic_age_group_code,
                          age_paired$susp_age_group_code,
                          paired = TRUE, alternative = "two.sided",
                          exact = FALSE)

# Kruskal-Wallis — both age codes across boroughs
age_long <- df %>%
  filter(vic_age_group_code > 0, boro_nm != "NULL") %>%
  select(boro_nm, vic_age_group_code)
t9_kw <- kruskal.test(vic_age_group_code ~ boro_nm, data = age_long)

cat("Test 9 — Age Group Analysis\n")
cat("  Sign Test: stat =", t9_sign_stat, "| p =",
    formatC(t9_sign_p, format = "e", digits = 4), "\n")
cat("  Wilcoxon Signed Rank:", "\n"); print(t9_wilcox)
cat("  Kruskal-Wallis (vic_age across boroughs):\n"); print(t9_kw)

test9_result <- data.frame(
  Test = "Sign + Wilcoxon + KW (Victim vs Suspect Age Groups)",
  Statistic = t9_sign_stat,
  P_Value   = min(t9_sign_p, t9_wilcox$p.value, t9_kw$p.value),
  Decision  = ifelse(t9_sign_p < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0"),
  Interpretation = paste(
    "Victim and suspect age group distributions",
    ifelse(t9_sign_p < 0.05, "differ significantly (Sign p < 0.05).",
           "do not significantly differ.")
  ),
  stringsAsFactors = FALSE
)
```

## Test 10 — Mann-Whitney + KW + KS (KY_CD across Bronx vs Queens)

```{r test10-multi}
bronx_ky <- na.omit(as.numeric(df$ky_cd[df$boro_nm == "BRONX"]))
qns_ky   <- na.omit(as.numeric(df$ky_cd[df$boro_nm == "QUEENS"]))

t10_mw <- wilcox.test(bronx_ky, qns_ky, alternative = "two.sided")
t10_ks <- ks.test(bronx_ky, qns_ky)

ky_all <- df %>%
  filter(boro_nm != "NULL", !is.na(ky_cd), ky_cd != "NULL") %>%
  mutate(ky_cd = as.numeric(ky_cd)) %>%
  filter(!is.na(ky_cd))
t10_kw <- kruskal.test(ky_cd ~ boro_nm, data = ky_all)

cat("Test 10 — KY_CD: Bronx vs Queens\n")
cat("  Mann-Whitney U:\n"); print(t10_mw)
cat("  KS Test:\n");        print(t10_ks)
cat("  Kruskal-Wallis (all boroughs):\n"); print(t10_kw)

test10_result <- data.frame(
  Test = "Mann-Whitney + KW + KS (KY_CD: Bronx vs Queens)",
  Statistic = round(as.numeric(t10_mw$statistic), 4),
  P_Value   = round(t10_mw$p.value, 6),
  Decision  = ifelse(t10_mw$p.value < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0"),
  Interpretation = paste(
    "Offense type (KY_CD) distributions",
    ifelse(t10_mw$p.value < 0.05,
           "differ significantly between Bronx and Queens.",
           "do not significantly differ between Bronx and Queens.")
  ),
  stringsAsFactors = FALSE
)
```

## Test 11 — Spearman + KS (Borough Code vs Offense Severity KY_CD)

```{r test11-spearman-ks}
boro_enc <- c("MANHATTAN" = 1, "BROOKLYN" = 2, "QUEENS" = 3,
              "BRONX" = 4, "STATEN ISLAND" = 5)
bks_df <- df %>%
  filter(boro_nm %in% names(boro_enc), !is.na(ky_cd), ky_cd != "NULL") %>%
  mutate(boro_code = boro_enc[boro_nm],
         ky_cd     = as.numeric(ky_cd)) %>%
  filter(!is.na(ky_cd))

t11_sp <- cor.test(bks_df$boro_code, bks_df$ky_cd,
                   method = "spearman", exact = FALSE)

# KS: Manhattan vs Bronx on KY_CD
mnh_ky_ks <- na.omit(as.numeric(df$ky_cd[df$boro_nm == "MANHATTAN"]))
bnx_ky_ks <- na.omit(as.numeric(df$ky_cd[df$boro_nm == "BRONX"]))
t11_ks    <- ks.test(mnh_ky_ks, bnx_ky_ks)

cat("Test 11 — Borough Code vs Offense KY_CD\n")
cat("  Spearman:\n"); print(t11_sp)
cat("  KS (Manhattan vs Bronx):\n"); print(t11_ks)

test11_result <- extract_test(
  "Spearman + KS (Borough Code vs Offense KY_CD)",
  t11_sp,
  paste("Borough predicts offense severity with ρ =",
        round(t11_sp$estimate, 3))
)
```

## Test 12 — Permutation Test (Median ADDR_PCT_CD: Manhattan vs Brooklyn)

**H₀:** Median ADDR_PCT_CD is equal in Manhattan and Brooklyn under resampling

```{r test12-permutation-median}
set.seed(42)
N_PERM2 <- 2000

obs_diff_med <- abs(median(mnh_pct) - median(bkn_pct))
combined_pct <- c(mnh_pct, bkn_pct)
n_mnh2       <- length(mnh_pct)

perm_meds <- replicate(N_PERM2, {
  s <- sample(combined_pct)
  abs(median(s[1:n_mnh2]) - median(s[(n_mnh2 + 1):length(s)]))
})

t12_p <- mean(perm_meds >= obs_diff_med)

cat("Permutation Test — Median ADDR_PCT_CD: Manhattan vs Brooklyn\n")
cat("  Observed |median diff|:", obs_diff_med, "\n")
cat("  P-value:", round(t12_p, 4), "\n")
cat("  Decision:", ifelse(t12_p < 0.05, "Reject H0", "Fail to Reject H0"), "\n")

test12_result <- data.frame(
  Test = "Permutation Test (Median ADDR_PCT_CD: Manhattan vs Brooklyn)",
  Statistic = obs_diff_med,
  P_Value   = round(t12_p, 4),
  Decision  = ifelse(t12_p < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0"),
  Interpretation = paste(
    "Resampling confirms: precinct code medians",
    ifelse(t12_p < 0.05,
           "differ significantly between Manhattan and Brooklyn.",
           "do not significantly differ between boroughs.")
  ),
  stringsAsFactors = FALSE
)
```

---

# Summary of All 12 Test Results

```{r results-summary}
all_results <- dplyr::bind_rows(
  test1_result,  test2_result,  test3_result,  test4_result,
  test5_result,  test6_result,  test7_result,  test8_result,
  test9_result,  test10_result, test11_result, test12_result
)

all_results$Test_Num <- 1:12
all_results <- all_results[, c("Test_Num", "Test", "Statistic",
                                "P_Value", "Decision", "Interpretation")]

kable(all_results,
      caption = "Assaults & Thefts — All 12 Nonparametric Test Results",
      col.names = c("#", "Test Name", "Statistic", "P-Value",
                    "Decision (α=0.05)", "Interpretation"))

# Count significant results
n_sig   <- sum(grepl("Reject", all_results$Decision) &
               !grepl("Fail", all_results$Decision))
n_total <- nrow(all_results)
cat("\nSignificant results:", n_sig, "of", n_total,
    "(", round(n_sig / n_total * 100, 1), "% )\n")

# Summary bar chart
sig_df <- all_results %>%
  mutate(
    Significant = ifelse(grepl("Reject", Decision) & !grepl("Fail", Decision),
                         "Significant (Reject H0)",
                         "Not Significant (Fail to Reject)")
  )
ggplot(sig_df, aes(x = factor(Test_Num), y = 1, fill = Significant)) +
  geom_col(width = 0.6) +
  scale_fill_manual(
    values = c("Significant (Reject H0)"  = "#10B981",
               "Not Significant (Fail to Reject)" = "#EF4444"),
    name = "Result"
  ) +
  labs(
    title    = "Assaults & Thefts — Nonparametric Test Results (12 Tests)",
    subtitle = paste(n_sig, "of", n_total, "tests significant at α = 0.05"),
    x = "Test Number", y = ""
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title   = element_text(face = "bold"),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid   = element_blank(),
    legend.position = "top"
  )

# Save results
results_path <- file.path(output_dir, "Assaults_Thefts_Test_Results.csv")
write.csv(all_results, results_path, row.names = FALSE)
cat("Results saved to:", results_path, "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
