---
title: "Nonparametric Analysis – NYC Fire Incidents (2015–2024)"
author: "Dharani Bhumireddy"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    df_print: kable
    number_sections: true
  html_document:
    toc: true
    toc_float: true
    theme: flatly
header-includes:
  - \usepackage{fontspec}
  - \setmainfont{Times New Roman}
  - \usepackage{booktabs}
  - \usepackage{longtable}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 10,
  fig.height = 6,
  dpi     = 150
)
```

# Overview

This R Markdown file performs the full nonparametric analysis for the
**NYC Fire Incident Dispatch** dataset (FDNY, 2015–2024).

- **Source:** NYC Open Data — Fire Incident Dispatch Data
- **Records:** ~6,054,083
- **Columns:** 29
- **Period:** 2015–2024

**10 nonparametric tests are executed in this file:**

| # | Test | Research Question |
|---|---|---|
| 1 | Sign Test | Is median incident response time significantly different from zero? |
| 2 | Wilcoxon Signed Rank | Do dispatch and incident response times differ significantly? |
| 3 | Wilcoxon Rank Sum | Do top-two ZIP codes differ in response times? |
| 4 | Kruskal-Wallis | Are response times different across City Council Districts? |
| 5 | Kolmogorov-Smirnov | Do response time distributions differ between top-two ZIP codes? |
| 6 | Spearman Correlation | Is dispatch time monotonically associated with incident response? |
| 7 | Kendall's Tau | Is there concordance between travel time and response time? |
| 8 | Spearman Correlation | Is travel time strongly associated with incident response time? |
| 9 | Friedman Test | Do median performance measures differ across years for each borough? |
| 10 | Runs Test | Does the sequence of fire incidents occur randomly over time? |

---

# Package Installation and Loading

```{r packages}
required_pkgs <- c(
  "dplyr", "tidyr", "stringr", "lubridate", "purrr",
  "readr", "ggplot2", "tseries", "coin", "knitr",
  "scales", "RColorBrewer", "gridExtra"
)
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[, "Package"])]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cran.r-project.org")
invisible(lapply(required_pkgs, library, character.only = TRUE))
set.seed(42)
```

---

# Data File Paths

```{r file-paths}
# ── UPDATE THESE PATHS ────────────────────────────────────────────────────
# Download from: https://data.cityofnewyork.us/Public-Safety/Fire-Incident-Dispatch-Data/8m42-w767

paths <- c(
  "data/NYC_FIRE_INCIDENTS/NYC_1_FIRE_INCIDENTS_RAW.csv",
  "data/NYC_FIRE_INCIDENTS/NYC_2_FIRE_INCIDENTS_RAW.csv",
  "data/NYC_FIRE_INCIDENTS/NYC_3_FIRE_INCIDENTS_RAW.csv"
)

output_dir <- "outputs/fire_incidents"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

---

# Helper Functions

```{r helper-functions}
# ── Read and clean one Fire Incidents file ─────────────────────────────────
safe_read_fire <- function(path) {
  df <- read.csv(path, stringsAsFactors = FALSE)
  colnames(df) <- toupper(trimws(colnames(df)))
  colnames(df) <- gsub("[[:space:]]+", "_", colnames(df))

  # Parse INCIDENT_DATETIME
  if ("INCIDENT_DATETIME" %in% colnames(df)) {
    df$INCIDENT_DATETIME <- suppressWarnings(
      lubridate::parse_date_time(
        df$INCIDENT_DATETIME,
        orders = c("ymd HMS", "ymd HM", "ymd",
                   "mdy HMS", "mdy HM", "mdy",
                   "dmy HMS", "dmy HM"),
        tz     = "America/New_York",
        quiet  = TRUE
      )
    )
  }
  df
}

# ── Prepare cleaned fire incidents dataframe ───────────────────────────────
prep_fire <- function(df) {
  valid_boroughs <- c("MANHATTAN", "BROOKLYN", "QUEENS", "BRONX", "STATEN ISLAND")
  df$BOROUGH <- toupper(trimws(df$BOROUGH))
  df <- df[df$BOROUGH %in% valid_boroughs & !is.na(df$BOROUGH), ]

  # Drop rows with missing datetime
  df <- df[!is.na(df$INCIDENT_DATETIME), ]

  # Numeric response time columns — must be positive
  time_cols <- c(
    "INCIDENT_RESPONSE_SECONDS_QY",
    "DISPATCH_RESPONSE_SECONDS_QY",
    "TRAVEL_TIME_SECONDS_QY"
  )
  for (cn in time_cols) {
    if (cn %in% names(df)) {
      df[[cn]] <- suppressWarnings(as.numeric(df[[cn]]))
      df        <- df[!is.na(df[[cn]]) & df[[cn]] > 0, ]
    }
  }

  # Add minutes columns for readability
  df$INCIDENT_RESPONSE_MINUTES  <- df$INCIDENT_RESPONSE_SECONDS_QY  / 60
  df$DISPATCH_RESPONSE_MINUTES  <- df$DISPATCH_RESPONSE_SECONDS_QY  / 60
  df$TRAVEL_TIME_MINUTES        <- df$TRAVEL_TIME_SECONDS_QY        / 60

  # Numeric quantity columns
  qty_cols <- c("ENGINES_ASSIGNED_QUANTITY", "LADDERS_ASSIGNED_QUANTITY",
                "OTHER_UNITS_ASSIGNED_QUANTITY")
  for (cn in qty_cols) {
    if (cn %in% names(df)) {
      df[[cn]] <- suppressWarnings(as.numeric(df[[cn]]))
      df[[cn]][is.na(df[[cn]])] <- 0
    }
  }

  # Temporal features
  df$YEAR        <- lubridate::year(df$INCIDENT_DATETIME)
  df$MONTH       <- lubridate::month(df$INCIDENT_DATETIME)
  df$HOUR        <- lubridate::hour(df$INCIDENT_DATETIME)
  df$DAY_OF_WEEK <- lubridate::wday(df$INCIDENT_DATETIME, label = TRUE)

  # CITY_COUNCIL_DIST — numeric
  if ("CITY_COUNCIL_DIST" %in% names(df)) {
    df$CITY_COUNCIL_DIST <- suppressWarnings(as.numeric(df$CITY_COUNCIL_DIST))
  }

  # Remove outside period
  df <- df[!is.na(df$YEAR) & df$YEAR >= 2015 & df$YEAR <= 2024, ]
  df
}

# ── Extract htest summary ─────────────────────────────────────────────────
extract_test <- function(test_name, result, interpretation) {
  if (is.null(result) || !inherits(result, "htest")) {
    return(data.frame(Test = test_name, Statistic = NA_real_,
                      P_Value = NA_real_, Decision = "NA",
                      Interpretation = interpretation,
                      stringsAsFactors = FALSE))
  }
  p_val    <- result$p.value
  decision <- ifelse(p_val < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0")
  data.frame(Test = test_name,
             Statistic = round(as.numeric(result$statistic), 4),
             P_Value   = round(p_val, 6),
             Decision  = decision,
             Interpretation = interpretation,
             stringsAsFactors = FALSE)
}
```

---

# Data Loading and Cleaning

```{r load-data}
cat("Loading fire incident files...\n")
fire_list <- lapply(paths, function(p) {
  cat("  Processing:", basename(p), "\n")
  df <- safe_read_fire(p)
  prep_fire(df)
})

df <- dplyr::bind_rows(fire_list)

cat("\nTotal records:", format(nrow(df), big.mark = ","), "\n")
cat("Columns:", ncol(df), "\n")
cat("Years:", min(df$YEAR), "–", max(df$YEAR), "\n")
cat("Boroughs:", paste(sort(unique(df$BOROUGH)), collapse = ", "), "\n")

# Response time summary (in minutes)
cat("\nResponse time summary (minutes):\n")
summary(df[, c("INCIDENT_RESPONSE_MINUTES",
               "DISPATCH_RESPONSE_MINUTES",
               "TRAVEL_TIME_MINUTES")])
```

---

# Save Cleaned Data

```{r save-cleaned}
clean_path <- file.path(output_dir, "fire_incidents_cleaned.csv")
write.csv(df, clean_path, row.names = FALSE)
cat("Cleaned data saved to:", clean_path, "\n")
```

---

# Exploratory Data Analysis

```{r eda-setup}
BOROUGH_COLORS <- c(
  "MANHATTAN"     = "#4F46E5",
  "BROOKLYN"      = "#EF4444",
  "QUEENS"        = "#10B981",
  "BRONX"         = "#F59E0B",
  "STATEN ISLAND" = "#8B5CF6"
)
```

```{r eda-response-time-borough}
resp_boro <- df %>%
  group_by(BOROUGH) %>%
  summarise(
    total_incidents    = n(),
    median_response    = round(median(INCIDENT_RESPONSE_MINUTES, na.rm = TRUE), 2),
    mean_response      = round(mean(INCIDENT_RESPONSE_MINUTES, na.rm = TRUE), 2),
    median_dispatch    = round(median(DISPATCH_RESPONSE_MINUTES, na.rm = TRUE), 2),
    median_travel      = round(median(TRAVEL_TIME_MINUTES, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(median_response)

kable(resp_boro,
      caption = "Fire Incident Response Time by Borough",
      col.names = c("Borough", "Total Incidents", "Median Response (min)",
                    "Mean Response (min)", "Median Dispatch (min)",
                    "Median Travel (min)"))

ggplot(resp_boro, aes(x = reorder(BOROUGH, median_response),
                      y = median_response, fill = BOROUGH)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = paste0(median_response, " min")),
            hjust = -0.1, size = 3.5, fontface = "bold") +
  geom_hline(yintercept = 6, linetype = "dashed",
             color = "#EF4444", linewidth = 1) +
  annotate("text", x = 0.6, y = 6.2,
           label = "FDNY 6-min target", color = "#EF4444",
           hjust = 0, size = 3.2) +
  scale_fill_manual(values = BOROUGH_COLORS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.20))) +
  coord_flip() +
  labs(
    title    = "Median Fire Incident Response Time by Borough",
    subtitle = "Kruskal-Wallis: Response times differ significantly across boroughs",
    x = "Borough", y = "Median Response Time (minutes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r eda-response-components}
resp_comp <- df %>%
  group_by(BOROUGH) %>%
  summarise(
    Dispatch = median(DISPATCH_RESPONSE_MINUTES, na.rm = TRUE),
    Travel   = median(TRAVEL_TIME_MINUTES, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  tidyr::pivot_longer(cols = c(Dispatch, Travel),
                      names_to = "Component", values_to = "minutes")

ggplot(resp_comp, aes(x = BOROUGH, y = minutes, fill = Component)) +
  geom_col(position = "stack", width = 0.6) +
  scale_fill_manual(values = c("Dispatch" = "#4F46E5", "Travel" = "#EF4444"),
                    name = "Component") +
  labs(
    title    = "Response Time Components by Borough (Median, Minutes)",
    subtitle = "Wilcoxon Signed Rank: Dispatch ≠ Total response (p<0.001)",
    x = "Borough", y = "Median Minutes"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 15, hjust = 1))
```

```{r eda-yearly-trend}
yr_resp <- df %>%
  group_by(YEAR, BOROUGH) %>%
  summarise(median_resp = median(INCIDENT_RESPONSE_MINUTES, na.rm = TRUE),
            .groups = "drop")

ggplot(yr_resp, aes(x = YEAR, y = median_resp,
                    color = BOROUGH, group = BOROUGH)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = BOROUGH_COLORS, name = "Borough") +
  scale_x_continuous(breaks = 2015:2024) +
  labs(
    title    = "Fire Incident Response Time Trend by Year and Borough",
    subtitle = "Friedman Test: Year-wise variation analysis",
    x = "Year", y = "Median Response Time (minutes)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r eda-incident-type}
inc_type_cnt <- df %>%
  filter(!is.na(INCIDENT_CLASSIFICATION_GROUP),
         INCIDENT_CLASSIFICATION_GROUP != "") %>%
  count(INCIDENT_CLASSIFICATION_GROUP, name = "n") %>%
  mutate(pct = round(n / sum(n) * 100, 1)) %>%
  arrange(desc(n))

kable(inc_type_cnt, caption = "Incident Classification Distribution",
      col.names = c("Classification", "Count", "% of Total"))
```

```{r eda-dispatch-scatter}
samp_df <- df %>%
  sample_n(min(3000, n())) %>%
  select(DISPATCH_RESPONSE_MINUTES, INCIDENT_RESPONSE_MINUTES,
         TRAVEL_TIME_MINUTES, BOROUGH)

ggplot(samp_df, aes(x = DISPATCH_RESPONSE_MINUTES,
                    y = INCIDENT_RESPONSE_MINUTES,
                    color = BOROUGH)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black",
              linewidth = 1.2, linetype = "dashed") +
  scale_color_manual(values = BOROUGH_COLORS, name = "Borough") +
  scale_x_continuous(limits = c(0, 30)) +
  scale_y_continuous(limits = c(0, 60)) +
  labs(
    title    = "Dispatch Time vs Total Incident Response Time",
    subtitle = "Spearman ρ and Kendall τ: Strong positive association",
    x = "Dispatch Response (minutes)",
    y = "Total Incident Response (minutes)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Nonparametric Tests

## Test 1 — Sign Test (Median Dispatch Time > 0)

**H₀:** Median dispatch response time equals zero
**H₁:** Median dispatch response time is significantly greater than zero

```{r test1-sign}
disp_vals  <- na.omit(df$DISPATCH_RESPONSE_SECONDS_QY)
signs_pos  <- sum(disp_vals > 0)
signs_neg  <- sum(disp_vals < 0)
n_sign     <- signs_pos + signs_neg
t1_stat    <- min(signs_pos, signs_neg)
t1_p       <- 2 * pbinom(t1_stat, size = n_sign, prob = 0.5)

cat("Sign Test — Median dispatch response time vs 0\n")
cat("  Positive signs:", format(signs_pos, big.mark = ","), "\n")
cat("  Negative signs:", signs_neg, "\n")
cat("  Test stat (min):", t1_stat, "\n")
cat("  P-value:", formatC(t1_p, format = "e", digits = 4), "\n")
cat("  Decision:", ifelse(t1_p < 0.05, "Reject H0", "Fail to Reject H0"), "\n")

test1_result <- data.frame(
  Test = "Sign Test (Median dispatch response time > 0)",
  Statistic = t1_stat,
  P_Value   = t1_p,
  Decision  = ifelse(t1_p < 0.05, "Reject H0 (p<0.05)", "Fail to Reject H0"),
  Interpretation = paste(
    "Median dispatch response time",
    ifelse(t1_p < 0.05,
           "is significantly different from zero — FDNY dispatch latency confirmed.",
           "is not significantly different from zero.")
  ),
  stringsAsFactors = FALSE
)
```

## Test 2 — Wilcoxon Signed Rank (Dispatch vs Incident Response Time)

**H₀:** Dispatch time and total incident response time are equal
**H₁:** Significant paired difference exists

```{r test2-wilcoxon-paired}
paired_df <- df %>%
  select(DISPATCH_RESPONSE_SECONDS_QY, INCIDENT_RESPONSE_SECONDS_QY) %>%
  filter(complete.cases(.))

# Sample if large (for computational speed)
if (nrow(paired_df) > 50000) {
  paired_df <- paired_df %>% sample_n(50000)
}

t2 <- wilcox.test(
  paired_df$DISPATCH_RESPONSE_SECONDS_QY,
  paired_df$INCIDENT_RESPONSE_SECONDS_QY,
  paired      = TRUE,
  alternative = "two.sided",
  exact       = FALSE
)

cat("Wilcoxon Signed Rank — Dispatch vs Incident response time\n")
print(t2)

test2_result <- extract_test(
  "Wilcoxon Signed Rank (Dispatch vs Incident response time — paired)",
  t2,
  paste("Dispatch and total incident response times",
        ifelse(t2$p.value < 0.05,
               "are significantly different — travel time adds meaningful delay.",
               "are not significantly different."))
)
```

## Test 3 — Wilcoxon Rank Sum (Top 2 ZIP Codes — Response Times)

**H₀:** Response times are equal in the top 2 highest-response ZIP codes
**H₁:** Response times differ significantly between the top 2 ZIPs

```{r test3-wilcoxon-ranksum-zip}
zip_medians <- df %>%
  group_by(ZIPCODE) %>%
  summarise(med_resp = median(INCIDENT_RESPONSE_SECONDS_QY, na.rm = TRUE),
            n = n(), .groups = "drop") %>%
  filter(n >= 30) %>%
  slice_max(med_resp, n = 2)

cat("Top 2 ZIP codes by median response time:\n")
print(zip_medians)

zip1_data <- df$INCIDENT_RESPONSE_SECONDS_QY[df$ZIPCODE == zip_medians$ZIPCODE[1]]
zip2_data <- df$INCIDENT_RESPONSE_SECONDS_QY[df$ZIPCODE == zip_medians$ZIPCODE[2]]
zip1_data <- na.omit(zip1_data)
zip2_data <- na.omit(zip2_data)

t3 <- wilcox.test(zip1_data, zip2_data, alternative = "two.sided",
                  correct = TRUE)

cat("\nWilcoxon Rank Sum — Top 2 ZIP response times\n")
print(t3)

# Box plot comparison
zip_df <- data.frame(
  response = c(zip1_data / 60, zip2_data / 60),
  zip      = c(rep(as.character(zip_medians$ZIPCODE[1]), length(zip1_data)),
               rep(as.character(zip_medians$ZIPCODE[2]), length(zip2_data)))
)

ggplot(zip_df, aes(x = zip, y = response, fill = zip)) +
  geom_boxplot(alpha = 0.8, outlier.size = 0.5, show.legend = FALSE,
               notch = TRUE) +
  scale_fill_manual(values = c("#4F46E5", "#EF4444")) +
  scale_y_continuous(limits = c(0, 60)) +
  labs(
    title    = "Response Time Distribution: Top 2 ZIP Codes",
    subtitle = paste("Wilcoxon Rank Sum: W =", round(t3$statistic, 0),
                     "| p =", round(t3$p.value, 4)),
    x = "ZIP Code", y = "Response Time (minutes)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test3_result <- extract_test(
  "Wilcoxon Rank Sum (Top 2 ZIP codes — response time)",
  t3,
  paste("Response times",
        ifelse(t3$p.value < 0.05,
               "differ significantly between the top 2 ZIP codes.",
               "do not significantly differ between the top 2 ZIP codes."))
)
```

## Test 4 — Kruskal-Wallis (Response Times across City Council Districts)

**H₀:** Response times are equal across City Council Districts
**H₁:** At least one district has significantly different response times

```{r test4-kruskal-wallis-district}
dist_df <- df %>%
  filter(!is.na(CITY_COUNCIL_DIST), CITY_COUNCIL_DIST > 0) %>%
  mutate(DISTRICT = factor(CITY_COUNCIL_DIST))

# Use top 15 most frequent districts
top_districts <- dist_df %>%
  count(DISTRICT) %>%
  slice_max(n, n = 15) %>%
  pull(DISTRICT)

dist_subset <- dist_df %>%
  filter(DISTRICT %in% top_districts)

t4 <- kruskal.test(INCIDENT_RESPONSE_SECONDS_QY ~ DISTRICT,
                   data = dist_subset)

cat("Kruskal-Wallis — Response times across City Council Districts\n")
print(t4)

# Box plot
ggplot(dist_subset, aes(x = DISTRICT,
                         y = INCIDENT_RESPONSE_MINUTES)) +
  geom_boxplot(fill = "#10B981", alpha = 0.7, outlier.size = 0.3) +
  scale_y_continuous(limits = c(0, 30)) +
  labs(
    title    = "Response Time by City Council District (Top 15)",
    subtitle = paste("Kruskal-Wallis: χ² =", round(t4$statistic, 2),
                     "| p =", round(t4$p.value, 4)),
    x = "City Council District", y = "Response Time (minutes)"
  ) +
  theme_minimal(base_size = 10) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))

test4_result <- extract_test(
  "Kruskal-Wallis (Response times across City Council Districts)",
  t4,
  paste("Response times",
        ifelse(t4$p.value < 0.05,
               "vary significantly across City Council Districts.",
               "do not vary significantly across City Council Districts."))
)
```

## Test 5 — Kolmogorov-Smirnov (Response Time Distributions: Top 2 ZIPs)

**H₀:** Response time distributions have the same shape in top 2 ZIP codes
**H₁:** Distribution shapes differ

```{r test5-ks}
t5 <- ks.test(zip1_data, zip2_data)

cat("KS Test — Response time distributions: Top 2 ZIP codes\n")
print(t5)

# ECDF
zip_ecdf_df <- data.frame(
  resp = c(zip1_data / 60, zip2_data / 60),
  zip  = c(rep(as.character(zip_medians$ZIPCODE[1]), length(zip1_data)),
           rep(as.character(zip_medians$ZIPCODE[2]), length(zip2_data)))
)

ggplot(zip_ecdf_df, aes(x = resp, color = zip)) +
  stat_ecdf(linewidth = 1.2) +
  scale_color_manual(values = c("#4F46E5", "#EF4444"), name = "ZIP Code") +
  scale_x_continuous(limits = c(0, 40)) +
  labs(
    title    = "ECDF: Response Time Distribution — Top 2 ZIP Codes (KS Test)",
    subtitle = paste("KS D =", round(t5$statistic, 4),
                     "| p =", round(t5$p.value, 4)),
    x = "Response Time (minutes)", y = "Cumulative Probability"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test5_result <- extract_test(
  "KS Test (Response time distributions: Top 2 ZIP codes)",
  t5,
  paste("Response time distribution shapes",
        ifelse(t5$p.value < 0.05,
               "differ significantly between the top 2 ZIP codes.",
               "do not significantly differ between the top 2 ZIP codes."))
)
```

## Test 6 — Spearman Correlation (Dispatch Time vs Incident Response)

**H₀:** No monotonic association between dispatch time and total response time
**H₁:** Significant monotonic association exists

```{r test6-spearman-dispatch}
corr_df <- df %>%
  select(DISPATCH_RESPONSE_SECONDS_QY, INCIDENT_RESPONSE_SECONDS_QY) %>%
  filter(complete.cases(.)) %>%
  sample_n(min(20000, n()))

t6 <- cor.test(corr_df$DISPATCH_RESPONSE_SECONDS_QY,
               corr_df$INCIDENT_RESPONSE_SECONDS_QY,
               method = "spearman", exact = FALSE)

cat("Spearman Correlation — Dispatch vs Incident response\n")
print(t6)

test6_result <- extract_test(
  "Spearman Correlation (Dispatch time vs Incident response time)",
  t6,
  paste("Strong monotonic association ρ =", round(t6$estimate, 3),
        "— dispatch delays drive total response time.")
)
```

## Test 7 — Kendall's Tau (Travel Time vs Incident Response)

**H₀:** No concordance between travel time and total incident response time
**H₁:** Significant concordant relationship exists

```{r test7-kendall}
# Sample for Kendall's tau (computationally expensive)
kt_df <- df %>%
  select(TRAVEL_TIME_SECONDS_QY, INCIDENT_RESPONSE_SECONDS_QY) %>%
  filter(complete.cases(.)) %>%
  sample_n(min(5000, n()))

t7 <- cor.test(kt_df$TRAVEL_TIME_SECONDS_QY,
               kt_df$INCIDENT_RESPONSE_SECONDS_QY,
               method = "kendall", exact = FALSE)

cat("Kendall's Tau — Travel time vs Incident response\n")
print(t7)

test7_result <- extract_test(
  "Kendall's Tau (Travel time vs Incident response time)",
  t7,
  paste("Concordance τ =", round(t7$estimate, 3),
        "— travel time is concordantly associated with total response time.")
)
```

## Test 8 — Spearman Correlation (Travel Time vs Incident Response)

**H₀:** No rank-based monotonic relationship between travel time and response time
**H₁:** Significant rank association exists

```{r test8-spearman-travel}
sp_df <- df %>%
  select(TRAVEL_TIME_SECONDS_QY, INCIDENT_RESPONSE_SECONDS_QY) %>%
  filter(complete.cases(.)) %>%
  sample_n(min(20000, n()))

t8 <- cor.test(sp_df$TRAVEL_TIME_SECONDS_QY,
               sp_df$INCIDENT_RESPONSE_SECONDS_QY,
               method = "spearman", exact = FALSE)

cat("Spearman Correlation — Travel time vs Incident response\n")
print(t8)

# Scatter plot
ggplot(sp_df %>% sample_n(min(2000, nrow(.))),
       aes(x = TRAVEL_TIME_SECONDS_QY / 60,
           y = INCIDENT_RESPONSE_SECONDS_QY / 60)) +
  geom_point(alpha = 0.25, color = "#10B981", size = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "#EF4444",
              linewidth = 1.2) +
  scale_x_continuous(limits = c(0, 30)) +
  scale_y_continuous(limits = c(0, 60)) +
  labs(
    title    = "Travel Time vs Total Incident Response Time",
    subtitle = paste("Spearman ρ =", round(t8$estimate, 4),
                     "| p =", formatC(t8$p.value, format = "e", digits = 3)),
    x = "Travel Time (minutes)", y = "Total Response (minutes)"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test8_result <- extract_test(
  "Spearman Correlation (Travel time vs Incident response time)",
  t8,
  paste("Rank-based association ρ =", round(t8$estimate, 3),
        "confirms travel time as a key driver of total response time.")
)
```

## Test 9 — Friedman Test (Year-wise Borough Response Time Patterns)

**H₀:** Median response times do not differ across years for each borough
**H₁:** Significant year × borough interaction in response times

```{r test9-friedman}
frd_fire <- df %>%
  group_by(BOROUGH, YEAR) %>%
  summarise(median_resp = median(INCIDENT_RESPONSE_SECONDS_QY, na.rm = TRUE),
            .groups = "drop") %>%
  tidyr::pivot_wider(names_from = YEAR, values_from = median_resp)

frd_matrix <- frd_fire %>%
  select(-BOROUGH) %>%
  as.matrix()
rownames(frd_matrix) <- frd_fire$BOROUGH
frd_complete <- frd_matrix[complete.cases(frd_matrix), ]

cat("Friedman matrix:\n")
print(round(frd_complete / 60, 2))  # print in minutes

if (nrow(frd_complete) >= 2 && ncol(frd_complete) >= 3) {
  t9 <- friedman.test(frd_complete)
  cat("\nFriedman Test — Year × Borough response times\n")
  print(t9)
  test9_result <- extract_test(
    "Friedman Test (Year-wise borough response time patterns)",
    t9,
    paste("Response time patterns",
          ifelse(t9$p.value < 0.05,
                 "differ significantly across years and boroughs.",
                 "do not significantly differ across years within boroughs."))
  )
} else {
  cat("Insufficient data for Friedman Test\n")
  test9_result <- data.frame(
    Test = "Friedman Test (Year-wise borough response time patterns)",
    Statistic = NA, P_Value = NA,
    Decision = "Insufficient data",
    Interpretation = "Need >= 2 boroughs and >= 3 years of complete data.",
    stringsAsFactors = FALSE
  )
}
```

## Test 10 — Runs Test (Temporal Randomness of Fire Incidents)

**H₀:** The sequence of monthly fire incidents is random
**H₁:** Non-random temporal patterns exist in fire incident frequency

```{r test10-runs}
monthly_fire <- df %>%
  count(YEAR, MONTH, name = "count") %>%
  arrange(YEAR, MONTH) %>%
  pull(count)

med_fire   <- median(monthly_fire, na.rm = TRUE)
binary_seq <- factor(ifelse(monthly_fire >= med_fire, "above", "below"))

t10 <- tseries::runs.test(binary_seq)

cat("Runs Test — Temporal randomness of monthly fire incidents\n")
print(t10)

ggplot(data.frame(idx = seq_along(monthly_fire), count = monthly_fire),
       aes(x = idx, y = count)) +
  geom_line(color = "#10B981", linewidth = 0.9) +
  geom_hline(yintercept = med_fire, linetype = "dashed",
             color = "#EF4444", linewidth = 1) +
  annotate("text", x = 1, y = med_fire * 1.03,
           label = "Median", color = "#EF4444", hjust = 0) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Monthly Fire Incidents — Temporal Sequence",
    subtitle = paste("Runs Test: z =", round(t10$statistic, 4),
                     "| p =", round(t10$p.value, 4)),
    x = "Month Index (2015–2024)", y = "Number of Incidents"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"))

test10_result <- extract_test(
  "Runs Test (Temporal randomness of fire incidents)",
  t10,
  paste("Fire incident sequence is",
        ifelse(t10$p.value < 0.05,
               "non-random — seasonal and temporal patterns detected.",
               "consistent with randomness — no strong temporal structure."))
)
```

---

# Summary of All 10 Test Results

```{r results-summary}
all_results <- dplyr::bind_rows(
  test1_result, test2_result, test3_result, test4_result, test5_result,
  test6_result, test7_result, test8_result, test9_result, test10_result
)
all_results$Test_Num <- 1:10
all_results <- all_results[, c("Test_Num", "Test", "Statistic",
                                "P_Value", "Decision", "Interpretation")]

kable(all_results,
      caption = "Fire Incidents — All 10 Nonparametric Test Results",
      col.names = c("#", "Test Name", "Statistic", "P-Value",
                    "Decision (α=0.05)", "Interpretation"))

n_sig <- sum(grepl("Reject", all_results$Decision) &
             !grepl("Fail", all_results$Decision))
cat("\nSignificant results:", n_sig, "of 10\n")

results_path <- file.path(output_dir, "Fire_Incidents_Test_Results.csv")
write.csv(all_results, results_path, row.names = FALSE)
cat("Results saved to:", results_path, "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
